Relative parabolic factorizations of Chevalley groups

### 2015-04-21 paper.tex
Added lemma on relative asr condition

For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda(\pi)$ the set of weights of $\pi$.
It will be convenient to use standard numbering for elements of $\Lambda(\pi)$ in the case when $\Phi$ is classical and $\pi$ is a vector representation (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, l+1 & \text{in the case} & \Phi =\rA_l, \\
  1,2,\ldots l, 0, -l,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_l, \\
  1,2,\ldots l, -l,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_l, \rD_l. \\
\end{array}$$
In particular, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ and so on.

\subsection{Elementary subgroup}

Recall that elementary root unipotents $t_{\alpha}(\xi)$, $x\in R$, $\alpha\in\Phi$ satisfy the following set of relations called Steinberg relations.
\begin{equation}\label{rel:add} t_\alpha(s) t_\alpha(t) = t_\alpha(s+t), \end{equation} 
\begin{equation}\label{rel:CCF} [t_\alpha(s),  t_\beta(t)] = \left(\prod\limits t_{i\alpha + j\beta}(N_{\alpha,\beta, i, j}s^i t^j)\right),\quad \alpha\neq-\beta, N_{\alpha, \beta, i, j}\in\mathbb{Z}.\end{equation}
The indices $i$, $j$ in the above formula range over all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha, \beta, i, j}$ are called {\it structure constants} of the Chevalley group $\G(\Phi,R)$ and depend only on $\Phi$.

For $\xi\in R^*$ set $w_\alpha(\xi) := x_\alpha(\xi) x_{-\alpha}(-\xi^{-1}) x_{\alpha}(\xi).$
If $\rk(\Phi)\geq 2$ the following relation is a consequence of \ref{rel:add}--\ref{rel:CCF}:
\begin{equation}\label{rel:R3} w_\alpha(\xi) t_{\beta}(s) w_\alpha(\xi)^{-1} =
t_{\sigma\ssub{\alpha}\beta} (\eta_{\alpha, \beta}\xi^{-\langle\beta,\alpha \rangle} s),\ \xi,s\in R. \end{equation}

Let $\Pi=\{\alpha_1, \ldots, \alpha_l\}$ be a basis of $\Phi$. Denote by $\Phi^+$ and $\Phi^-$ subsets of positive and negative roots with respect to $\Pi$.
Denote by $m_i(\alpha)$ the coefficient of $\beta$ in the expansion of $\alpha$ in $\Pi$, i.\,e. $\alpha = \sum\limits_{i=1}^n m_i(\alpha) \alpha_i$.

>>>>

For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda(\pi)$ the set of weights of $\pi$.
It will be convenient to use standard numbering for weights of representation $\pi$ in the case when $\Phi$ is classical and $\pi$ is a vector representation (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, l+1 & \text{in the case} & \Phi =\rA_l, \\
  1,2,\ldots l, 0, -l,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_l, \\
  1,2,\ldots l, -l,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_l, \rD_l. \\
\end{array}$$
In particular, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ and so on.

\subsection{Elementary subgroup}
Recall that elementary root unipotents $t_{\alpha}(\xi)$, $x\in R$, $\alpha\in\Phi$ satisfy the following set of relations called Steinberg relations.
\begin{equation}\label{rel:add} t_\alpha(s) t_\alpha(t) = t_\alpha(s+t), \end{equation} 
\begin{equation}\label{rel:CCF} [t_\alpha(s),  t_\beta(t)] = \left(\prod\limits t_{i\alpha + j\beta}(N_{\alpha,\beta, i, j}s^i t^j)\right),\quad \alpha\neq-\beta, N_{\alpha, \beta, i, j}\in\mathbb{Z}.\end{equation}
The indices $i$, $j$ in the above formula range over all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha, \beta, i, j}$ are called {\it structure constants} of the Chevalley group $\G(\Phi,R)$ and depend only on $\Phi$.

For $\xi\in R^*$ set $w_\alpha(\xi) := x_\alpha(\xi) x_{-\alpha}(-\xi^{-1}) x_{\alpha}(\xi).$
If $\rk(\Phi)\geq 2$ the following relation is a consequence of \ref{rel:add}--\ref{rel:CCF}:
\begin{equation}\label{rel:R3} w_\alpha(\xi) t_{\beta}(s) w_\alpha(\xi)^{-1} =
t_{\sigma\ssub{\alpha}\beta} (\eta_{\alpha, \beta}\xi^{-\langle\beta,\alpha \rangle} s),\ \xi,s\in R. \end{equation}
where coefficients $\eta_{\alpha, \beta} = \pm 1$ (see~\cite[\S~13]{VP} for more details on $\eta_{\alpha, \beta}$).

\begin{rem} Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ of a classical group such that $\lambda_1-\lambda_2\in \Phi$.
In this situation it will be convenient for us to write $t_{\lambda_1,\lambda_2}(\xi)$ instead of $t_{\lambda_1-\lambda_2}(\xi)$.
In particular, for $\Phi=\rA_\ell$ we have $t_{1,2}(\xi)=t_{2-1}(\xi)=t_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = t_{\alpha_1}(\xi)$. \end{rem}

\subsection{Root systems}
Let $\Pi=\{\alpha_1, \ldots, \alpha_l\}$ be a basis of $\Phi$. Denote by $\Phi^+$ and $\Phi^-$ subsets of positive and negative roots with respect to $\Pi$.
Denote by $m_i(\alpha)$ the coefficient of $\beta$ in the expansion of $\alpha$ in $\Pi$, i.\,e. $\alpha = \sum\limits_{i=1}^n m_i(\alpha) \alpha_i$.

### 2015-04-21 stability-conditions.tex
Added lemma on relative asr condition

\subsection{Absolute stable rank}
For a row $a=(a_1,\ldots,a_n)\in{}^n\!R$ denote by $\mathfrak{J}(a)$ the intersection of all maximal ideals of $R$ containing $a_1,\ldots,a_n$.
It is easy to see that a row $a\in R^n$ is unimodular if and only if $\mathfrak{J}(a)=R$. 
Clearly, for any $g\in\GL(n,R)$ one has $\mathfrak{J}(a\cdot g)=\mathfrak{J}(a)$.
\begin{dfn}
We say that a pair $(R,I)$ satisfies $\ASR_n(R,I)$ if it satisfies $\SR_n(R,I)$ and, moreover, 
for any row $a=(a_1,\ldots,a_{n+1})\in I^n$ there exist $b_1,\ldots,b_n\in I$ such that
\[\mathfrak{J}(a_1,\ldots,a_{n+1})=\mathfrak{J}(a_1+b_1a_{n+1},\ldots,a_n+b_na_{n+1}).\]
It is easy to see that $\ASR_m(R,I)$ implies $\ASR_n(R,I)$ for any $n\geqslant m$. 
By definition, the \emph{absolute relative stable rank} $\asr(R,I)$ is the smallest natural $n$ such that $\ASR_n(R, I)$ holds. 
Similarly, we set $\asr(R,I)=\infty$ if no such $n$ exists.
When $I=R$ we call $\asr(R, R)$ simply the \emph{absolute stable rank} of $R$ and denote it by $\asr(R)$.
\end{dfn}
The following reformulation of the absolute stable rank condition was noted by A.~Bak.
\begin{lemma} For a commutative ring $R$ the following statements are equivalent:
\begin{enumerate}
 \item\label{asr_Stein} $R$ satisfies $\ASR_n(R)$;
 \item\label{asr_Bak} For any \emph{unimodular} row $(a_1,\ldots,a_n,b,d)\in\Ums(n+2,R)$ there exist $c_1,\ldots,c_n\in R$ such that $(a_1+c_1d,\ldots,a_n+c_nd,b+td)$ is unimodular for all $t\in R$.
\end{enumerate}
\end{lemma}
\begin{proof}
Assume first that $\ASR_n(R)$ holds.
Then for the row $(a_1,\ldots,a_n,d)$ there exist $c_1,\ldots,c_n$ such that $\mathfrak{J}(a_1,\ldots,a_n,d)=\mathfrak{J}(a_1+c_1d,\ldots,a_n+c_nd)$. 
Therefore $a_1+c_1d,\ldots,a_n+c_nd$ together with $b$ generate the unit ideal.
Of course, one could replace $b$ with $b+td$ from the very start.

@@

The following lemma illustrates a typical application of stability conditions.
\begin{lemma} \label{lemma:uraction} 
Let $\Phi=\rA_\ell, \rD_\ell$. Denote by $\pi$ standard vector representation of $\G(\Phi, R)$ on $V=R^n$, $n=\ell+1,2\ell$.
Assume that one of the following conditions holds:
\begin{itemize}
 \item $\Phi=\rA_\ell$, $\Gamma=\{ k+1, \ldots, \ell+1\} \subset \Lambda(\pi)$ and $\sr(I)\leq k\leq \ell$;
 \item $\Phi=\rD_\ell$, $\Gamma=\{-l,\ldots, -2, -1\} \subset \Lambda(\pi)$ and $\asr(R, I)\leq \ell-1$. 
\end{itemize}
Then for any $I$-unimodular column $v\in \Ums(n, R, I)$ there exist $x\in \U(\Phi^+, I)$, $y\in \U(\Phi^-, I)$ such that $(yx \cdot v)_\lambda = 0$ for $\lambda\in \Gamma$.
\end{lemma}
\textbf{TODO: Write proof of this lemma}
%TODO:

>>>>

\subsection{Absolute stable rank}
For a row $a=(a_1,\ldots,a_n)\in{}^n\!R$ denote by $\mathfrak{J}(a)$ the intersection of all maximal ideals of $R$ containing $a_1,\ldots,a_n$.
It is easy to see that a row $a\in R^n$ is unimodular if and only if $\mathfrak{J}(a)=R$. 
Clearly, for any $g\in\GL(n,R)$ one has $\mathfrak{J}(a\cdot g)=\mathfrak{J}(a)$.
\begin{dfn}
We say that a pair $(R,I)$ satisfies $\ASR_n(R,I)$ if it satisfies $\SR_n(R,I)$ and, moreover, 
for any row $a=(a_1,\ldots,a_{n+1})\in I^n$ there exist $b_1,\ldots,b_n\in I$ such that
\[\mathfrak{J}(a_1,\ldots,a_{n+1})=\mathfrak{J}(a_1+b_1a_{n+1},\ldots,a_n+b_na_{n+1}).\]
It is easy to see that $\ASR_m(R,I)$ implies $\ASR_n(R,I)$ for any $n\geqslant m$. 
By definition, the \emph{absolute relative stable rank} $\asr(R,I)$ is the smallest natural $n$ such that $\ASR_n(R, I)$ holds. 
Similarly, we set $\asr(R,I)=\infty$ if no such $n$ exists.
When $I=R$ we call $\asr(R, R)$ simply the \emph{absolute stable rank} of $R$ and denote it by $\asr(R)$.
\end{dfn}
The following reformulation of the absolute stable rank condition was noted by A.~Bak.
\begin{lemma} For a commutative ring $R$ the following statements are equivalent:
\begin{enumerate}
 \item\label{asr_Stein} $R$ satisfies $\ASR_n(R, R)$;
 \item\label{asr_Bak} For any \emph{unimodular} row $(a_1,\ldots,a_n,b,d)\in\Ums(n+2,R)$ there exist $c_1,\ldots,c_n\in R$ such that $(a_1+c_1d,\ldots,a_n+c_nd,b+td)$ is unimodular for all $t\in R$.
\end{enumerate}
\end{lemma}
\begin{proof}
Assume first that $\ASR_n(R)$ holds.
Then for the row $(a_1,\ldots,a_n,d)$ there exist $c_1,\ldots,c_n$ such that $\mathfrak{J}(a_1,\ldots,a_n,d)=\mathfrak{J}(a_1+c_1d,\ldots,a_n+c_nd)$. 
Therefore $a_1+c_1d,\ldots,a_n+c_nd$ together with $b$ generate the unit ideal.
Of course, one could replace $b$ with $b+td$ from the very start.

@@

The next lemma illustrates a typical application of the absolute relative stability condition.
Let $v\in R^{2\ell}$ be a vector of the standard vector representation of $\G(\rD_\ell, R)$.
It will be convenient for us to denote by $v^+$ and $v^-$ the upper and lower halves of $v$ i.e. $v^+=(v_1,\ldots, v_\ell)^t$, $v^-=(v_{-\ell},\ldots, v_{-1})^t$.
\begin{lemma}\label{lemma:asrUnip}
 Let $\Phi=\rD_\ell$ and assume that $\asr(R, I)\leq \ell -1$. Then for any $I$-unimodular column $v=(v^+, v^-)^t\in\Ums(2\ell, R, I)$
 there exist $g\in\U(\Sigma^+_\ell, I)$ such that $(v\cdot g)^+ \in \Ums(\ell, R, I)$.
\end{lemma}
\begin{proof} Denote by $J$ the ideal of $R$ spanned by entries of $v^-$. Clearly, $J \leq I$.
By Proposition~\ref{prop:sr_properties} $\sr(I/J) \leq \ell-1$ and therefore
the elementary group $\E(\rA_{\ell-1}, R/J, I/J)$ acts transitively on $\Ums(\ell, R/J, I/J)$. 
This implies existence of $h\in \E(\Delta_\ell, R, I)$ such that for $v' = v\cdot h$ one has $(v')^+ \equiv e_1$ modulo $J$.

By hypothesis, we can find $c_2,\ldots, c_\ell\in I$ such that for $v''=v'\cdot \prod\limits_{i=2}^{\ell}t_{-1, -i}(c_i)$ one has
$\mathfrak{J}(v''_{-\ell},\ldots, v''_{-2}) = \mathfrak{J}(v'^-)$. Clearly, $v''_1$ is still congruent to $1$ modulo $J$, therefore 
$(v''_1, v''_{-\ell},\ldots, v''_{-1})$ is $I$-unimodular. Now, applying condition $\sr(I) \leq \ell-1$ once again we find
$d_1,d_3,\ldots, d_{\ell}\in I$ such that for $v'''=v''\cdot t_{-2,1}(d_1) \cdot \prod\limits_{i=3}^{\ell} t_{-2,-i}(d_i)$
the entries $(v'''_1, v'''_{-\ell},\ldots, v'''_{-3})$ form a $I$-unimodular column.

We can find $f_1, f_3,\ldots, f_\ell \in R$ such that $f_1v'''_1+\sum\limits_{i=3}^\ell f_i v'''_{-i} = 1$.
Set $\xi = v'''_1-v'''_2-1 \in I$, $v^4=v'''\cdot t_{1,2}(\xi f_1) \cdot \prod\limits_{i=3}^\ell t_{-i,2}(\xi i_k)$.
Clearly $v^4_1 = v'''_1$ and $v^4_2 = v'''_1-1$, therefore $v^{4+}$ is $I$-unimodular.
Summarizing the above, we have found $g\in \EP_\ell(R, I)$ such that $v^4=v\cdot g$
and assertion of the lemma immediately follows from Levi decomposition. \end{proof}

\begin{lemma} \label{lemma:uraction} 
Let $\Phi=\rA_\ell, \rD_\ell$. Denote by $\pi$ standard vector representation of $\G(\Phi, R)$ on $V=R^n$, $n=\ell+1,2\ell$.
Assume that one of the following conditions holds:
\begin{itemize}
 \item $\Phi=\rA_\ell$, $\Gamma=\{ k+1, \ldots, \ell+1\} \subset \Lambda(\pi)$ and $\sr(I)\leq k\leq \ell$;
 \item $\Phi=\rD_\ell$, $\Gamma=\{-l,\ldots, -2, -1\} \subset \Lambda(\pi)$ and $\asr(R, I)\leq \ell-1$. 
\end{itemize}
Then for any $I$-unimodular column $v\in \Ums(n, R, I)$ there exist $x\in \U(\Phi^+, I)$, $y\in \U(\Phi^-, I)$ such that $(yx \cdot v)_\lambda = 0$ for $\lambda\in \Gamma$.
\end{lemma}
\textbf{TODO: Write proof of this lemma}
%TODO:
